---
title: extracting train/test from data set within RDBMS
author: Ben Bolker
---

**Goal**: extract train/test data set from a data set stored in a data base.

**desiderata**:

- efficient
- clean/pretty syntax (I prefer `dbplyr`/tidyverse because I'm more familiar with it ...)
- platform-independent (not all databases have the same back-end capabilities, although if something works in (R)SQLite then it probably works pretty broadly ...)
- reproducible

```{r pkgs,message=FALSE}
library("RSQLite")
library("dbplyr")
library("dplyr")
library("rbenchmark")
```

"How do I efficiently/robustly sample rows from a database?" is a *very* frequent question, see e.g. [here](http://jan.kneschke.de/projects/mysql/order-by-rand/) or [here](http://mysql.rjweb.org/doc.php/random), or [this tutorial](http://www.mysqltutorial.org/select-random-records-database-table.aspx), or this [SO question](https://stackoverflow.com/questions/580639/how-to-randomly-select-rows-in-sql/1209946#1209946) ...

```{r defs}
nr <- 1e5
nc <- 900
```

Make up a data set to play with:

```{r testdat,cache=TRUE}
dd <- data.frame(id=1:nr,matrix(rnorm(nr*nc),nrow=nr))
dim(dd)
## clean up
unlink("test.sqlite")
unlink("test.sqlite-journal")
test.sqlite = dbConnect(SQLite(),dbname="test.sqlite")
dbWriteTable(test.sqlite, "test", dd, overwrite=TRUE)
```

Reconnect to data frame (if necessary because previous chunk is cached):
```{r reconnect}
test.sqlite = dbConnect(SQLite(),dbname="test.sqlite")
```

Note that SQLite's `random()` function (which is its version of `RAND()`, more common in SQL-based systems) **cannot be seeded**
(see [here](https://stackoverflow.com/questions/2171578/seeding-sqlite-random) and [here](https://stackoverflow.com/questions/24256258/order-by-random-with-seed-in-sqlite)),
so this solution is not reproducible. Ugh.

```{r queries}
queries <- c(nonrand='SELECT * from test limit 1000',
     rand1='SELECT * from test order by random() limit 1000',
     rand2='SELECT * FROM test WHERE random()<\
(SELECT ((1000/COUNT(*))*10) FROM test) LIMIT 1000',
     rand3='SELECT t.* FROM test AS t JOIN \
(SELECT ROUND(random() * (SELECT MAX(id) FROM test)) AS id) AS \
x WHERE t.id >= x.id LIMIT 1000'
)
```

Make sure all queries return the right dimensions ...
```{r test_q,cache=TRUE}
for (q in queries) {
    ## print(q)
    if (!identical(qdim <- dim(dbGetQuery(test.sqlite,q)),c(1000L,as.integer(nc+1)))) {
        cat(q,"\n")
        print(qdim)
    }
}
```
(this chunk appears to be not entirely reproducible/deterministic - usually works, but fails *sometimes*?)

We can use these queries in a tidyverse workflow by using `tbl(sql(.))`
(note we need to use `[[]]` in extracting elements from a named vector ...)

```{r test_tidy}
stopifnot(identical(train_dat0 <- (test.sqlite %>% tbl(sql(queries[[1]])) %>% collect()),
                    as_tibble(dbGetQuery(test.sqlite,queries[[1]]))))
```

```{r run_bench,cache=TRUE}
b1 <- benchmark(
    dbGetQuery(test.sqlite,queries[["nonrand"]]),
    dbGetQuery(test.sqlite,queries[["rand1"]]),
    dbGetQuery(test.sqlite,queries[["rand2"]]),
    dbGetQuery(test.sqlite,queries[["rand3"]]),
    test.sqlite %>% tbl(sql(queries[["rand3"]])) %>% collect(),
    replications=50,
    columns=c('test', 'elapsed', 'replications', 'relative')
)
```

```{r print_bench}
print(b1)
```

The crude method (`ORDER by random()`) is actually fastest for this small database; reading the various SQL info suggests that it may not scale very well.

How about the possibly more sensible, reproducible method of
pulling the id values out first? This also makes it relatively
easy to get train and test sets by using `inner_join()` vs.
`anti_join()` ...

```{r tidyverse,cache=TRUE}
## extract ID column
all_ids <- (test.sqlite
    %>% tbl("test")
    %>% select(id)
    %>% collect()
)
## generate sample IDs for training set
## might want to compute ntot <- nrow(all_ids); s <- round(ntot/2)
train_ids <- (all_ids
    %>% pull(id)
    %>% sample(size=1000)
)
## inner join to extract only elements from training set;
## copy=TRUE copies the train IDs back into the database
## where they can be joined 
train_dat <- (test.sqlite
    %>% tbl("test")
    %>% inner_join(tibble(id=train_ids),copy=TRUE,by="id")
    %>% collect()
)
## anti-join to extract only elements NOT in training set
test_dat <-  (test.sqlite
    %>% tbl("test")
    %>% anti_join(tibble(id=train_ids),copy=TRUE,by="id")
    %>% collect()
)
```


Results are *not* the same as before because different random numbers
are used: the best we can do is check dimensions.

```{r test_tidy2}
all.equal(dim(train_dat0), dim(train_dat))
```

Not sure about speed ... what's a fair comparison?

```{r leftovers, echo=FALSE, eval=FALSE}
dataframe = "lending-club-loan-data/database.sqlite"
loan.sqlite = dbConnect(SQLite(),dbname=dataframe)
loan.tbl <- tbl(loan.sqlite,"loan")
## count total number of rows
ntot <- loan.tbl %>% count() %>% collect() %>% pull(n)
print(ntot)
set.seed(1)
train = sample(ntot,floor(ntot/2))
small <- dbGetQuery(loan.sqlite,'select * from loan limit 10000')
## smtrain <- dbGetQuery(loan.sqlite,'select * from loan order by random() limit 1000')
pop = dbGetQuery(loan.sqlite,'select count(*) from loan')
loan.tbl 
## inefficient but not necessarily a better way to do it?
loantrain.new <- (loan.tbl
    %>% filter(row_number() %in% train)
)

count_NAs <- (loantrain.new
    %>% summarise_all(funs(sum(is.na(.))))
    %>% collect()
)
        
    %>% select(loantrain,-c(index,
                            url,
                            desc,
                            id,
                            verification_status_joint,
                            pymnt_plan,
                            application_type,member_id,
                            inq_last_12m,
                            total_cu_tl,
                            inq_fi,
                            all_util,
                            max_bal_bc,
                            open_rv_24m,
                            open_rv_12m,
                            il_util,
                            total_bal_il,
                            mths_since_rcnt_il,
                            open_il_24m,
                            open_il_12m,
                            open_il_6m,
                            open_acc_6m,
                            dti_joint,
                            annual_inc_joint,
                            policy_code
                            ))


```
